\textls[-0]{\thetitle}

### 2023-04-01 metatheory-condensed.tex
Adding list of figures.
Adding List of Abbrev.
List of abbreviation added

We contribute \TT, a core type theory that extends
%\underline{f}a\underline{m}i\underline{l}y polymorphism
%\underline{M}artin-\underline{L}öf dependent \underline{t}ype \underline{t}heory (MLTT)~\cite{martin1982constructive}
\citeauthor{martin1982constructive} dependent type theory (MLTT)
with facilities to express family polymorphism
while maintaining consistency and canonicity.
%\NA{maybe you can give an overview of the salient points of the design and what they're trying to achieve? What is the core aspect modeled by the theory?}
%

@@

Our formulation of W-types differs from previous ones in that it is
straightforward to identify constructors from what we call \emph{W-type signatures}.
%
A signature $\goodWSig{\Gm}{\wsig}{n}$ is
composed of $n$ pairs of types (\ruleref{wsig/empty} and \ruleref{wsig/add}),
%\Zhixuan{Page 14: I don't see the necessity of introducing "enhanced W-types". Since there are already boolean types and Σ types, Fin types can be defined. Using Fin, ordinary W types suffice to define trees with n constructors. So it might be clearer to just define 'enhanced W-types' as syntactic sugar.}\YZreply{I concur.}\EDJreply{This following sentence has been replied to Zhixuan and he doesn't has further response: 1. We have to have enhanced W-type at FMLTT, that's for sure. Otherwise we (a). are too far away from the surface syntax because W-type itself only has one constructor all the time (b) this suggested encoding or syntactic sugar, since it interact with other connective like linkage (the Wrec is the interaction), so Wrec part is mostly unchanged  (c) since they will look the same, the encoding itself will look like an extra equality judgement in the syntax... which is not helpful for the complexity at all ... and even more complex  \\
%2. so to me the problem you guys really have is around the content in the appendix -- I think Zhixuan mean he doesn't think this is more expressive. But my point is, it is hard to really write out the above syntactic sugar, formally. I think the issue here is similar to other reviewer saying "syntax translation". It is very hard to write out the encoding of Fin already, and then encode it again into W-type. (Just ask my agda mechanization). Of course, some hand-waiving can be done. For example, the part Zhixuan says I think it okay to hand-wave. But the destruction part -- the part Wrec interact with this encoding, I don't believe handwaving is enough. \\
%3. But how do I say this to the audience? I mean the main reason I didn't choose syntactic translation/syntax sugar is basically ``doing encoding is harder than directly working in ambient logic.'' }
%
each modeling a constructor of the inductive type.
The $i$-th pair, projected from the signature~$\wsig$ using the forms
$\goodType{\Gm}{\wsigproj{i}{1}{\wsig}}{}$ and
$\goodTerm{\Gm}{\wsigproj{i}{2}{\wsig}}{\wsigproj{i}{1}{\wsig} \to \cU}$,
defines the $i$-th constructor.
That is, given two arguments $\goodTerm{\Gm}{t_1}{\wsigproj{i}{1}{\wsig}}$ and
$\goodTerm{\Gm, x : \El{\wsigproj{i}{2}{\wsig}(t_1)} }{t_2}{\El{\wcode{\wsig}}}$,
one can construct a term $\Wsup{i}{\wsig}{t_1}{x.t_2}$ of the W-type \El{\wcode{\wsig}} (\ruleref{tm/wsup}).
$\wcode{\wsig}$ gives the code of the
W-type (\ruleref{tm/w}).

@@

\definecolor{A-color}{HTML}{E7EFC5}
\definecolor{T-color}{HTML}{FFF1D0}
\definecolor{t-color}{HTML}{FAE0E4}
\definecolor{P-color}{HTML}{BFEFFF}

@@

The definitive version containing all the rules in \TT is stated in
\ifreport \cref{sec:complete-fmltt}.
\else a technical appendix.
\fi
This definitive version uses de Bruijn indices and explicit substitution.
An effort to mechanize the formalization is underway. It follows \citet{altkap2016}
in using quotient inductive-inductive types (QIITs) as 
the logical framework to state intrinsically typed syntax (and also
in using Agda's \lsti{REWRITE} pragma to replace the limited uses of equality
reflection, as Agda does not support QIITs natively).
%in the extensional meta\-logic.

>>>>

We contribute \ac{FMLTT}, a core type theory that extends
%\underline{f}a\underline{m}i\underline{l}y polymorphism
%\underline{M}artin-\underline{L}öf dependent \underline{t}ype \underline{t}heory (MLTT)~\cite{martin1982constructive}
\citeauthor{martin1982constructive} dependent type theory (MLTT)
with facilities to express family polymorphism
while maintaining consistency and canonicity.
%\NA{maybe you can give an overview of the salient points of the design and what they're trying to achieve? What is the core aspect modeled by the theory?}
%

@@

Our formulation of W-types differs from previous ones in that it is
straightforward to identify constructors from what we call \emph{W-type signatures}.
%
A signature $\goodWSig{\Gm}{\wsig}{n}$ is
composed of $n$ pairs of types (\ruleref{wsig/empty} and \ruleref{wsig/add}),
%\Zhixuan{Page 14: I don't see the necessity of introducing "enhanced W-types". Since there are already boolean types and Σ types, Fin types can be defined. Using Fin, ordinary W types suffice to define trees with n constructors. So it might be clearer to just define 'enhanced W-types' as syntactic sugar.}\YZreply{I concur.}\EDJreply{This following sentence has been replied to Zhixuan and he doesn't has further response: 1. We have to have enhanced W-type at \TT, that's for sure. Otherwise we (a). are too far away from the surface syntax because W-type itself only has one constructor all the time (b) this suggested encoding or syntactic sugar, since it interact with other connective like linkage (the Wrec is the interaction), so Wrec part is mostly unchanged  (c) since they will look the same, the encoding itself will look like an extra equality judgement in the syntax... which is not helpful for the complexity at all ... and even more complex  \\
%2. so to me the problem you guys really have is around the content in the appendix -- I think Zhixuan mean he doesn't think this is more expressive. But my point is, it is hard to really write out the above syntactic sugar, formally. I think the issue here is similar to other reviewer saying "syntax translation". It is very hard to write out the encoding of Fin already, and then encode it again into W-type. (Just ask my agda mechanization). Of course, some hand-waiving can be done. For example, the part Zhixuan says I think it okay to hand-wave. But the destruction part -- the part Wrec interact with this encoding, I don't believe handwaving is enough. \\
%3. But how do I say this to the audience? I mean the main reason I didn't choose syntactic translation/syntax sugar is basically ``doing encoding is harder than directly working in ambient logic.'' }
%
each modeling a constructor of the inductive type.
The $i$-th pair, projected from the signature~$\wsig$ using the forms
$\goodType{\Gm}{\wsigproj{i}{1}{\wsig}}{}$ and
$\goodTerm{\Gm}{\wsigproj{i}{2}{\wsig}}{\wsigproj{i}{1}{\wsig} \to \cU}$,
defines the $i$-th constructor.
That is, given two arguments $\goodTerm{\Gm}{t_1}{\wsigproj{i}{1}{\wsig}}$ and
$\goodTerm{\Gm, x : \El{\wsigproj{i}{2}{\wsig}(t_1)} }{t_2}{\El{\wcode{\wsig}}}$,
one can construct a term $\Wsup{i}{\wsig}{t_1}{x.t_2}$ of the W-type \El{\wcode{\wsig}} (\ruleref{tm/wsup}).
$\wcode{\wsig}$ gives the code of the
W-type (\ruleref{tm/w}).

@@

@@

The definitive version containing all the rules in \TT is stated in
\ifreport \cref{sec:complete-fmltt}.
\else a technical appendix.
\fi
This definitive version uses de Bruijn indices and explicit substitution.
An effort to mechanize the formalization is underway. It follows \citet{altkap2016}
in using \ac{QIIT} as 
the logical framework to state intrinsically typed syntax (and also
in using Agda's \lsti{REWRITE} pragma to replace the limited uses of equality
reflection, as Agda does not support QIITs natively).
%in the extensional meta\-logic.

### 2023-04-01 related-work.tex
fix
List of abbreviation added

\citet{delaware2011} engineer product lines of theorems and proofs built
from feature modules. Feature composition is manual, which seems to have
motivated later approaches based on \emph{data types à la carte} (DTC)~\cite{dtc}.
%\rB{ define abbreviation DTC at first use?}

%DTC is as a source of inspiration for may approaches
%\cite{delaware2013,delaware2013b,schwaab2013modular,keuchel2013generic,forsta2020}.
The original DTC encoding requires type-level general recursion that
fails the strict positivity check imposed by proof assistants including Coq and Agda.
\citet{delaware2013} introduce \emph{meta-theory à la carte} (MTC):
it overcomes the problem by using Church encodings for
data types and using Mendler-style folds for evaluation,
though it requires \lsti{Set} impredicativity.
Feature composition is automated through heavy use of type classes.
The framework is implemented as a Coq library.
\citet{schwaab2013modular} adapt DTC to Agda by considering a restricted
class of functors that admit least fixed points.
\citet{keuchel2013generic} use datatype-generic programming techniques
for the underlying representation of type-level fixed points and
avoid \lsti{Set} impredicativity.
%\Citet{vdr2022intrinsically} mechanize intrinsically-typed
%target language based on container and signature \cite{altenkirch2015container}. 
%Thanks to the nature of intrinsically-typed abstract syntax trees (ASTs) and Agda's advanced unification mechanism, the authors were able to eliminate a significant amount of code that dealt with "impossible" cases.
%\YZ{What can we say about Intrinsically-Typed Definitional Interpreters a La Carte?}\EDJreply{My understanding is,
% it is really an intrinsic typed version of MTC. It uses different set of tools (like containers (a la altenkirch, can be considered as an alternative of W-type) and stuff) but it is still a design pattern paper.
% But good news is, container is closer to Wtype(inductive type) than the endofunctor encoding (classic MTC approach)}\EDJreply{I also add one sentence after Coq a la carte. You can also check... I will add bib later. Interestingly, Intrinsic type doesn't refer Coq a la carte.}

@@

Pumpkin Pi~\cite{ringer2021pumpkin} is a Coq plug-in that helps
repair proofs broken by changes in type definitions.
Its decompiler from proof terms to proof scripts prioritizes suggesting
useful tactics over soundness.
While Pumpkin Pi focuses on refactoring existing proofs in response to
changed definitions,
our solution can be viewed as an effort to preempt refactoring by
enabling the programmer to write code that has built-in hooks for
future extensions.

>>>>

\citet{delaware2011} engineer product lines of theorems and proofs built
from feature modules. Feature composition is manual, which seems to have
motivated later approaches based on \ac{DTC}~\cite{dtc}.
%\rB{ define abbreviation DTC at first use?}

%DTC is as a source of inspiration for may approaches
%\cite{delaware2013,delaware2013b,schwaab2013modular,keuchel2013generic,forsta2020}.
The original DTC encoding requires type-level general recursion that
fails the strict positivity check imposed by proof assistants including Coq and Agda.
\citet{delaware2013} introduce \ac{MTC}:
it overcomes the problem by using Church encodings for
data types and using Mendler-style folds for evaluation,
though it requires \lsti{Set} impredicativity.
Feature composition is automated through heavy use of type classes.
The framework is implemented as a Coq library.
\citet{schwaab2013modular} adapt DTC to Agda by considering a restricted
class of functors that admit least fixed points.
\citet{keuchel2013generic} use datatype-generic programming techniques
for the underlying representation of type-level fixed points and
avoid \lsti{Set} impredicativity.
%\Citet{vdr2022intrinsically} mechanize intrinsically-typed
%target language based on container and signature \cite{altenkirch2015container}. 
%Thanks to the nature of intrinsically-typed abstract syntax trees (ASTs) and Agda's advanced unification mechanism, the authors were able to eliminate a significant amount of code that dealt with "impossible" cases.
%\YZ{What can we say about Intrinsically-Typed Definitional Interpreters a La Carte?}\EDJreply{My understanding is,
% it is really an intrinsic typed version of MTC. It uses different set of tools (like containers (a la altenkirch, can be considered as an alternative of W-type) and stuff) but it is still a design pattern paper.
% But good news is, container is closer to Wtype(inductive type) than the endofunctor encoding (classic MTC approach)}\EDJreply{I also add one sentence after Coq a la carte. You can also check... I will add bib later. Interestingly, Intrinsic type doesn't refer Coq a la carte.}

@@

Pumpkin Pi~\cite{ringer2021pumpkin} is a Coq plugin that helps
repair proofs broken by changes in type definitions.
Its decompiler from proof terms to proof scripts prioritizes suggesting
useful tactics over soundness.
While Pumpkin Pi focuses on refactoring existing proofs in response to
changed definitions,
our solution can be viewed as an effort to preempt refactoring by
enabling the programmer to write code that has built-in hooks for
future extensions.
